Practical Optional Types for Clojure

### 2015-10-16 intro.tex
need to cut 1.5 pages
1/2 page to go

We synthesize these lessons to present \emph{Typed Clojure}, an
optional type system for Clojure. 
%
Clojure is a dynamically
typed language in the Lisp family built to run on the Java Virtual
Machine (JVM) which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via a
standard library of immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for parallelism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, it now has an extensive base of existing untyped
programs, whose developers can now benefit from Typed Clojure. As
a result, Typed Clojure is used in industry, experience we discuss
in this paper.

@@

\begin{figure}[t!]
\inputminted[firstline=5]{clojure}{code/demo/src/demo/ex-intro.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

@@

But multimethods are not the only Clojure feature crucial to type
checking existing programs. As a language built on the Java Virtual
Machine, Clojure provides flexible and transparent access to existing
Java libraries, and \textbf{Clojure/Java interoperation} is found in almost
every significant Clojure code base. Typed Clojure therefore builds in
an understanding of the Java type system and handles interoperation
appropriately. Notably, Typed Clojure avoids conflating Java's
\texttt{null} with all other types, leading to automatic type-enforced
absence of null-pointer exceptions.

A simple example of these features in combination is given in
figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
on the \texttt{class} of the argument---when it is a \texttt{String},
the first method implementation is called, for a Java \texttt{File}, the
second. The method for \texttt{String} uses Java interoperation
to call a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
method for \texttt{File} invokes
the \texttt{.getName} method with a non-nil target, returning a nilable
type. Typed Clojure uses type information to resolve JVM method or constructor overloads,
avoiding expensive runtime reflective calls.

Finally, when not working with existing Java classes and objects, the
most common Clojure data structure is the dictionary, a
high-performance immutable table which Clojure programmers repurpose
for all manner of data types. Simply treating them as uniformly-typed
key-value mappings would be entirely insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous map} type, in which mandatory, optional, and
absent keys can be specified.

While these features may seem disparate, they are unified in important
ways. First, they all leverage the type system mechanisms that Typed
Clojure inherits from Typed Racket---multimethods when using 
dispatch via predicates, Java interoperation for handling
\texttt{null} tests, and heterogenous maps using union types and
reasoning about subcomponents of data. Second, and more significantly,
they are the crucial features for handling Clojure code in
practice. Typed Clojure's use in real Clojure deployments would not be
possible without effective handling of these three Clojure features.

@@

\begin{enumerate}
  \item We motivate and describe  Typed Clojure, an optional
    type system for Clojure that understands existing Clojure idioms.
  \item We present a sound formal model for the three crucial type
    system features that Typed Clojure relies on: multi-methods, Java
    interoperability, and heterogenous maps.
  \item We evaluate the use of Typed Clojure features on existing
    Typed Clojure code, including both open source and in-house systems.
\end{enumerate}

@@

\noindent
 The remainder of this paper begins with an example-driven
 presentation of the main type system features in
 \secref{sec:overview}. We then incrementally present a core calculus
 for Typed Clojure covering all of these features together in
 \secref{sec:formal} and prove type soundness
 (\secref{sec:metatheory}). We then discuss the full implementation of
 Typed Clojure, dubbed \coretyped{}, which extends the formal model in
 many ways, and empirical analysis of significant code bases written
 in \coretyped{} in \secref{sec:experience}. Finally, we discuss
 related work and conclude.

>>>>

We synthesize these lessons to present \emph{Typed Clojure}, an
optional type system for Clojure. 
%
Clojure is a dynamically
typed language in the Lisp family---built on the Java Virtual
Machine (JVM)---which has recently gained popularity as an alternative
JVM language.  It offers the flexibility of a Lisp dialect, including
macros, emphasizes a functional style via
immutable data structures, and provides
interoperability with existing Java code, allowing programmers to use
existing Java libraries without leaving Clojure.
%
Since its initial release in 2007, Clojure has been widely adopted for
``backend'' development in places where its support for parallelism,
functional programming, and Lisp-influenced abstraction is desired on
the JVM. As a result, it has an extensive base of existing untyped
programs, whose developers can benefit from Typed Clojure. As
a result, Typed Clojure is used in industry, experience we discuss
in this paper.

@@

\begin{figure}[t!]
\inputminted[firstline=15]{clojure}{code/demo/src/demo/ex-intro.clj}
\caption{A simple Typed Clojure program}
\label{fig:ex1}
\end{figure}

@@

But multimethods are not the only Clojure feature crucial to type
checking existing programs. As a language built on the Java Virtual
Machine, Clojure provides flexible and transparent access to existing
Java libraries, and \textbf{Clojure/Java interoperation} is found in almost
every significant Clojure code base. Typed Clojure therefore builds in
an understanding of the Java type system and handles interoperation
appropriately. Notably, \texttt{null} is a distinct type in Typed Clojure,
leading to automatic type-enforced
absence of null-pointer exceptions.

An example of these features is given in
figure~\ref{fig:ex1}. Here, the \texttt{pname} multimethod dispatches
on the \texttt{class} of the argument---for \texttt{String}s,
the first method implementation is called, for \texttt{File}s, the
second. The \texttt{String} method calls
a \texttt{File} constructor, returning a non-nil \texttt{File} instance---the 
\texttt{.getName} method 
on \texttt{File} requires a non-nil target, returning a nilable
type.  Typed Clojure offers an opt-in mode that
resolves JVM overloading, avoiding expensive runtime reflective calls.

Finally, high-performance immutable dictionaries
are the most common Clojure data structure,
which programmers repurpose liberally.
Simply treating them as uniformly-typed
key-value mappings would be insufficient for existing
programs and programming styles. Instead, Typed Clojure provides a
flexible \textbf{heterogenous map} type, in which specific entries can be specified.

While these features may seem disparate, they are unified in important
ways. First, they leverage the type system mechanisms
inherited from Typed Racket---multimethods when using 
dispatch via predicates, Java interoperation for handling
\texttt{null} tests, and heterogenous maps using union types and
reasoning about subcomponents of data. Second,
they are crucial features for handling Clojure code in
practice. Typed Clojure's use in real Clojure deployments would not be
possible without effective handling of these three Clojure features.

@@

\begin{enumerate}
  \item We motivate and describe  Typed Clojure, an optional
    type system for Clojure that understands existing Clojure idioms.
  \item We present a sound formal model for three crucial type
    system features: multi-methods, Java
    interoperability, and heterogenous maps.
  \item We evaluate the use of Typed Clojure features on existing
    Typed Clojure code, including both open source and in-house systems.
\end{enumerate}

@@

\noindent
 The remainder of this paper begins with an example-driven
 presentation of the main type system features in
 \secref{sec:overview}. We then incrementally present a core calculus
 for Typed Clojure covering all of these features together in
 \secref{sec:formal} and prove type soundness
 (\secref{sec:metatheory}). We then 
 %discuss the full implementation of
 %Typed Clojure, \coretyped{}, which extends the formal model in many ways, 
 present an empirical analysis of significant code bases written
 in the full implementation of Typed Clojure---\coretyped{}---in \secref{sec:experience}. 
 Finally, we discuss related work and conclude.

### 2015-10-16 overview.tex
1/2 page to go

Here is a simple program in Typed Clojure.
%
We define \clj{greet} as a one-argument function
taking and returning a string.

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n] (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Strings are accepted, but providing \clj{nil} (Clojure's name for Java's \java{null})
is a static type error---\clj{nil} is not a string in Typed Clojure.
%
\begin{minted}{clj}
(greet nil) ; Type Error: Expected Str, given nil
\end{minted}

\paragraph{Unions} We can make the annotation more permissive with \emph{ad-hoc unions}
to allow \clj{nil}.
All Clojure values are true except \clj{nil} and \clj{false}, so a \clj{nil} test
amounts to a test on the value.

\begin{minted}{clj}
(ann  greet-nil [(U nil Str) -> Str])
(defn greet-nil [n] (str "Hello" (when n (str ", " n)) "!"))
(greet-nil "Donald") ; :- Str
;=> "Hello, Donald!"
(greet-nil nil)      ; :- Str
;=> "Hello!"
\end{minted}

Typed Clojure guarantees that well-typed code cannot
dereference the null-pointer.
This is especially important for Clojure programs---\clj{nil}
is treated like any other distinct datum in Clojure,  and its status as
one of only two false values means it is a common choice to indicate
``nothing'' or ``failure''.

\paragraph{Flow analysis} Typed Clojure uses occurrence typing~\cite{TF10} to
model type-based control flow.
In \clj{greetings}, a branch ensures \clj{repeat}
is never passed \clj{nil}.
%
\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, " (when i (apply str (repeat i "hello, "))) n "!"))
(greetings "Donald" 2)  ; :- Str
;=> "Hello, hello, hello, Donald!"
(greetings "Grace" nil) ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Removing the branch is a static type error---\clj{repeat} cannot be passed \clj{nil}.
%
\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]           ; Expected Int, given (U nil Int)
  (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
\end{minted}

@@

Clojure supports interoperability with Java, including the ability to
call constructors, invoke methods, and access fields.

The following Typed Clojure program constructs a new \clj{File}
instance and calls the \clj{getParent} method on the result,
returning a nullable string \clj{"a"}.

\begin{exmp}
\begin{minted}{clj}
(.getParent (new File "a/b"))  ; :- (U nil Str)
;=> "a"
\end{minted}
\label{example:getparent-direct-constructor}
\end{exmp}

Typed Clojure helps the Clojure compiler avoid expensive reflective 
calls,
however if a specific constructor, method, or field cannot be found based on the
static types of its arguments, a type error is thrown.

\begin{minted}{clj}
(fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent

\end{minted}

Function arguments default to \clj{Any}, the most permissive type. Ascribing
a parameter type allows Typed Clojure to find a specific method.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error: Cannot call instance method on nil.
\end{minted}
Since Java-level types do not provide information about nullability,
by default Typed Clojure conservatively assumes that Java method and
constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
\egref{example:getparent-direct-constructor},
 must be \emph{non-nullable}, but this can be
configured for  particular calls as needed.
The target of method invocations is always non-nullable.

Any Java method returning a reference can also return \java{null} ---
Typed Clojure rejects programs that assume otherwise.
%
\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
\end{minted}
%
%Typed Clojure and Java treat \java{null} differently.
%In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
%called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
%This means the Java static type \java{String} is equivalent to
%\clj{(U nil String)} in Typed Clojure.
%
%Reference types in Java are nullable, so to guarantee a method call does not
%leak \java{null} into a Typed Clojure program we
%must assume methods can return \clj{nil}.
%
In contrast, JVM invariants guarantee constructors return non-null.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
%
\begin{exmp}
\inputminted[firstline=15,lastline=15]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

@@

\paragraph{Class-based dispatch}
For class values, multimethods can choose methods based on subclassing
relationships.
%
Recall the multimethod from figure~\ref{fig:ex1}, reproduced here.
\begin{minted}{clj}
(ann pname [(U File String) -> (U nil String)])
(defmulti pname class)
(defmethod pname String [s] (pname (new File s)))
(defmethod pname File [f] (.getName f))
\end{minted}

%Its dispatching function is
%\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
%respectively.
\noindent
The dispatch function \clj{class}
%---associated at multimethod creation with \clj{defmulti}---
dictates 
whether the \clj{String} or \clj{File} method is chosen.
%---both installed via \clj{defmethod}
%
The multimethod dispatch rules use
\clj{isa?}, a hybrid predicate which is a subclassing check for classes and
an equality check for other values.

\begin{minted}{clojure}
(isa? (identity :en) :en) ;=> true
(isa? (identity :en) :fr) ;=> false
(isa? (class "STAINS/JELLY") String) ;=> true
(isa? (class "STAINS/JELLY") Object) ;=> true
(isa? (class (new File "JELLY")) String) ;=> false
\end{minted}
%
The current dispatch value and---in turn---each method's associated dispatch value
is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.

In our example,
  \clj{(pname "STAINS/JELLY")}
chooses the  method for \clj{String} because
\clj{(isa? (class "STAINS/JELLY") String)}
is true and
\clj{(isa? (class "STAINS/JELLY") File)}
is false. The \clj{String} method body
\clj{(pname (new File "STAINS/JELLY"))}
chooses the \clj{File} method for opposite reasons,
resulting in 
\begin{minted}{clojure}
(.getName (new File "STAINS/JELLY")) ;=> "JELLY" :- (U nil Str)
\end{minted}

@@

Keywords look themselves up in a map when called, or return \clj{nil}.
%
\begin{exmp}
\begin{minted}{clojure}
(def breakfast {:en "waffles" :fr "croissants"})
(:en breakfast)  ;=> "waffles" :- Str
(:bocce breakfast) ; nil :- nil
\end{minted}
\label{example:breakfastcomplete}
\end{exmp}

@@

HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
omits
information about absent keys, only providing information on
\clj{:mandatory} keys.
%
\begin{exmp}
\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})
(:bocce lunch) ; nil :- Any ; less accurate type
\end{minted}
\label{example:lunchpartial}
\end{exmp}
%(:en lunch)    ; :- Str
%;=> "muffin"
%(:fr lunch)    ; :- Str
%;=> "baguette"

@@

%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
If we forget to remove the unencrypted private key,
a type error is given, because \clj{EncKeyPair} is fully specified.
%
\begin{minted}{clojure}
(ann enc-keypair-bad [RawKeyPair -> EncKeyPair])
(defn enc-keypair-bad [{pk :priv :as kp}]
  (assoc kp :enc-priv (encrypt pk))) ; Type error: :pub disallowed
\end{minted}
The extra \clj{:priv} entry does not match \clj{EncKeyPair}, so a type error
is raised.

>>>>

Here is a simple one-argument function
%
\clj{greet} that takes and returns strings.

\begin{minted}{clj}
(ann  greet [Str -> Str])
(defn greet [n] (str "Hello, " n "!"))
(greet "Grace") ; :- Str
;=> "Hello, Grace!"
\end{minted}
%
Providing \clj{nil} (exactly Java's \java{null})
is a static type error---\clj{nil} is not a string.
%
\begin{minted}{clj}
(greet nil) ; Type Error: Expected Str, given nil
\end{minted}

\paragraph{Unions} To allow \clj{nil}, we use \emph{ad-hoc unions} (\clj{nil} and \clj{false}
are logically false).

\begin{minted}{clj}
(ann  greet-nil [(U nil Str) -> Str])
(defn greet-nil [n] (str "Hello" (when n (str ", " n)) "!"))
(greet-nil "Donald") ;=> "Hello, Donald!" :- Str 
(greet-nil nil)      ;=> "Hello!"         :- Str
\end{minted}

Typed Clojure prevents well-typed code from dereferencing \clj{nil}.
%This is important for Clojure programs---\clj{nil}
%is treated like any other distinct datum in Clojure.

\paragraph{Flow analysis} Occurrence typing~\cite{TF10}
models type-based control flow.
In \clj{greetings}, a branch ensures \clj{repeat}
is never passed \clj{nil}.
%
\begin{minted}{clj}
(ann  greetings [Str (U nil Int) -> Str])
(defn greetings [n i]
  (str "Hello, " (when i (apply str (repeat i "hello, "))) n "!"))
(greetings "Donald" 2)  ;=> "Hello, hello, hello, Donald!" :- Str
(greetings "Grace" nil) ;=> "Hello, Grace!"                :- Str
\end{minted}
%
Removing the branch is a static type error---\clj{repeat} cannot be passed \clj{nil}.
%
\begin{minted}{clj}
(ann  greetings-bad [Str (U nil Int) -> Str])
(defn greetings-bad [n i]           ; Expected Int, given (U nil Int)
  (str "Hello, " (apply str (repeat i "hello, ")) n "!"))
\end{minted}

@@

Clojure can interact with Java constructors, methods, and fields.
This program calls the \clj{getParent} on a constructed
\clj{File}
instance, inferred as a nullable string.

\begin{exmp}
\begin{minted}{clj}
(.getParent (new File "a/b"))  ;=> "a" :- (U nil Str)
\end{minted}
\label{example:getparent-direct-constructor}
\end{exmp}
%
Typed Clojure can optionally integrate with the Clojure compiler to avoid expensive reflective 
calls, however if a specific constructor, method, or field cannot be found based on the
static types of its arguments, a type error is thrown.
%
\begin{minted}{clj}
(fn [f] (.getParent f)) ; Type Error: Unresolved interop: getParent
\end{minted}
%
Function arguments default to \clj{Any}, the most permissive type. Ascribing
a parameter type allows Typed Clojure to find a specific method.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error: Cannot call instance method on nil.
\end{minted}
Since Java-level types lack nullability information,
Typed Clojure conservatively asserts Java methods and
constructor arguments cannot be passed \clj{nil}---this restriction can be
adjusted per method. Method targets can never be \clj{nil}.

Typed Clojure rejects programs that assume methods cannot return \clj{nil}.
%
\begin{minted}{clj}
(defn parent-bad-out [f :- File] :- Str
  (.getParent f)) ; Type Error: Expected Str, given (U nil Str).
\end{minted}
%
%Typed Clojure and Java treat \java{null} differently.
%In Clojure, where it is known as \clj{nil}, Typed Clojure assigns it an explicit type
%called \clj{nil}. In Java \java{null} is implicitly a member of any reference type.
%This means the Java static type \java{String} is equivalent to
%\clj{(U nil String)} in Typed Clojure.
%
%Reference types in Java are nullable, so to guarantee a method call does not
%leak \java{null} into a Typed Clojure program we
%must assume methods can return \clj{nil}.
%
In contrast, JVM invariants guarantee constructors return non-null.\footnote{\url{http://docs.oracle.com/javase/specs/jls/se7/html/jls-15.html#jls-15.9.4}}
%
\begin{exmp}
\inputminted[firstline=15,lastline=15]{clojure}{code/demo/src/demo/parent3.clj}
\end{exmp}

@@

\paragraph{Class-based dispatch}
For class values, multimethods can choose methods based on subclassing
relationships.
%
Recall the multimethod from \figref{fig:ex1}. %, reproduced here.
%\begin{minted}{clj}
%(ann pname [(U File String) -> (U nil String)])
%(defmulti pname class)
%(defmethod pname String [s] (pname (new File s)))
%(defmethod pname File [f] (.getName f))
%\end{minted}
%
%Its dispatching function is
%\clj{class}, with two methods associated with dispatch values \clj{java.lang.String} and \clj{java.io.File}
%respectively.
%\noindent
The dispatch function \clj{class}
%---associated at multimethod creation with \clj{defmulti}---
dictates 
whether the \clj{String} or \clj{File} method is chosen.
%---both installed via \clj{defmethod}
%
The multimethod dispatch rules use
\clj{isa?}, a hybrid predicate which is a subclassing check for classes and
an equality check for other values.

%(isa? (class "STAINS/JELLY") Object) ;=> true
%(isa? (identity :en) :fr) ;=> false
%(isa? (class (new File "JELLY")) String) ;=> false
\begin{minted}{clojure}
(isa? :en :en)       ;=> true
(isa? String Object) ;=> true
\end{minted}
%
The current dispatch value and---in turn---each method's associated dispatch value
is supplied to \clj{isa?}. If exactly one method returns true, it is chosen.
%
For example,
the call
  \clj{(pname "STAINS/JELLY")}
picks the \clj{String} method because \clj{(isa? String String)}
is true, and
\clj{(isa? String File)}
is not.
%---\clj{(class "STAINS/JELLY")}
%is \clj{String}.
%
%The \clj{String} method body
%\clj{(pname (new File "STAINS/JELLY"))}
%chooses the \clj{File} method for opposite reasons.

@@

Keywords look themselves up in a map when called, or return \clj{nil}.
%
\begin{exmp}
\begin{minted}{clojure}
(def breakfast {:en "waffles" :fr "croissants"})
(:en breakfast)    ;=> "waffles" :- Str
(:bocce breakfast) ;=> nil       :- nil
\end{minted}
\label{example:breakfastcomplete}
\end{exmp}

@@

HMap types default to  partial specification---\clj{:complete?} defaults to \clj{false}. The HMap shorthand 
\clj{'{:en Str :fr Str}}
omits
information about absent keys, only providing information on
\clj{:mandatory} keys.
%
\begin{exmp}
\begin{minted}{clojure}
(ann lunch '{:en Str :fr Str})
(def lunch {:en "muffin" :fr "baguette"})
(:bocce lunch) ;=> nil :- Any ; less accurate type
\end{minted}
\label{example:lunchpartial}
\end{exmp}
%(:en lunch)    ; :- Str
%;=> "muffin"
%(:fr lunch)    ; :- Str
%;=> "baguette"

@@

%\inputminted[firstline=10,lastline=22]{clojure}{code/demo/src/demo/key.clj}
As \clj{EncKeyPair} is fully specified, we must remove extra keys like \clj{:priv}.
%
\begin{minted}{clojure}
(defn enc-keypair-bad [{pk :priv :as kp}]
  (assoc kp :enc-priv (encrypt pk))) ; Type error: :priv disallowed
\end{minted}

### 2015-10-16 small.tex
New page for bib. Changes to sec 1 & 2
CircleCI data
add GT citations. better update figure and section 3.6
convert to ESOP format
fix esop formatting
need to cut 1.5 pages
1/2 page to go

\documentclass[preprint,9pt,twocolumn,nocopyrightspace,authoryear]{sigplanconf}

@@

\usepackage{common}
\usepackage{minted}
\include{bibliography.bib}
\usepackage{chngcntr}
\usepackage{url}

\newcounter{ex}
\newenvironment{exmp}{
\refstepcounter{ex}%
\par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
\\[-2.4em]\nopagebreak}
{\relax}

@@

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

@@

\authorinfo{}{}{\vspace*{-2mm}}
%\authorinfo{Ambrose Bonnaire-Sergeant}
%           {Indiana University}
%           {abonnair@indiana.edu}
%\authorinfo{Rowan Davies}
%           {University of Western Australia}
%           {rowan.davies@uwa.edu.au}
%\authorinfo{Sam Tobin-Hochstadt}
%           {Indiana University}
%           {samth@indiana.edu}

@@

\begin{abstract}
Typed Clojure is an optional type system for Clojure, a dynamic
language in the Lisp family that targets the JVM. Typed Clojure
enables Clojure programmers to gain greater confidence in the
correctness of their code via static type checking while remaining in
the Clojure world, and has acquired significant adoption in the
Clojure community. Typed Clojure's type
system repurposes Typed Racket's
\emph{occurrence typing}, an approach to statically reasoning about
predicate tests, and also includes several new type system features to
handle existing Clojure idioms.

@@

\input{acknowledgements}

@@

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

>>>>

\documentclass{llncs}

@@

\bibliographystyle{splncs03}

\usepackage{common}
\usepackage{minted}
\include{bibliography.bib}
\usepackage{chngcntr}
\usepackage{url}
\newcounter{ex}
\newenvironment{exmp}{
\refstepcounter{ex}%
\par\vspace{.8em}\hfill\framebox{\small Example~\theex}\nopagebreak
\\[-2.5em]\nopagebreak}
{\relax}

@@

%
%\setminted{%fontsize=\footnotesize
%}

@@

%\authorinfo{}{}{\vspace*{-2mm}}
\author{\ma{\text{Ambrose Bonnaire-Sergeant}^\dagger\text{, Rowan Davies*}, \text{Sam Tobin-Hochstadt}^\dagger}}
\institute{Indiana University$\dagger$; Omnia Team, Commonwealth Bank of Australia*}
%       {Indiana University}
%       {abonnair@indiana.edu}
%       {University of Western Australia}
%       {Rowan.Davies@cba.com.au}
%       {Indiana University}
%       {samth@indiana.edu}

@@

\begin{abstract}
Typed Clojure is an optional type system for Clojure, a dynamic
language in the Lisp family that targets the JVM. Typed Clojure
enables Clojure programmers to gain greater confidence in the
correctness of their code via static type checking while remaining in
the Clojure world, and has acquired significant adoption in the
Clojure community. Typed Clojure repurposes Typed Racket's
\emph{occurrence typing}, an approach to statically reasoning about
predicate tests, and also includes several new type system features to
handle existing Clojure idioms.

@@

%\input{acknowledgements}

@@

% The bibliography should be embedded for final submission.
\newpage
